% ============================================================
% A Machine-Checked Atlas of Riemann-Hypothesis Equivalences
% and No-Go Certificates
%
% Formal-methods preprint — 2026-06-16.
% Companion artifact: the Lean 4 library in this repository, `formal/`
% (128 source modules, axiom-clean, zero sorries).
% Compile: pdflatex or tectonic (no external .bib; thebibliography inline).
% Bibliography verified against primary sources (2026-06).
% ============================================================
\documentclass[11pt]{amsart}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{seqsplit} % allow long Lean identifiers to break across lines
\usepackage[hidelinks]{hyperref}
\raggedbottom

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Xifn}{\Xi}
\newcommand{\Gammar}{\Gamma_{\!\R}}
\newcommand{\code}[1]{\texttt{\seqsplit{#1}}}
\let\Re\relax % override LaTeX's built-in \Re (fraktur) so we can use "Re"
\DeclareMathOperator{\Re}{Re}
\DeclareMathOperator{\Spec}{Spec}

\begin{document}
% Many long Lean identifiers appear inline; allow looser spacing so \texttt
% runs do not overrun the margin.
\sloppy

\title[A machine-checked atlas of RH equivalences and no-go certificates]{A
machine-checked atlas of Riemann-Hypothesis equivalences and no-go
certificates}

% AUTHORSHIP NOTE (to be settled by M.~Haufschild before submission):
% author list, affiliation, and the framing of the AI contribution are the
% repository owner's call. The Lean development was produced by AI research
% agents ("Fable" = Claude, and GPT) under the author's direction, with the
% Lean kernel as final referee.
\author{Michael Haufschild}
\email{haufschildmichael@gmail.com}

\date{June 16, 2026}

\begin{abstract}
We describe \code{jensen-ladder}, a Lean~4 library that formalizes, against
\code{mathlib}, the \emph{logical geometry} of the Riemann Hypothesis (RH):
the network of equivalent reformulations and the structural obstructions that
any proof must respect. The library contains $128$ source modules; the full
\code{lake build} is green ($8603$ compiled jobs); the tree contains no
\code{sorry} and no added \code{axiom}, and the headline results depend only on
the three standard \code{mathlib} axioms \code{[propext, Classical.choice,
Quot.sound]}. Three things are mechanized. First, a short, unconditional
\emph{reduction spine} that proves \code{mathlib}'s official
\code{RiemannHypothesis} equivalent to the reality of the \emph{regular} zeros
of the completed zeta function in a symmetric variable, with the
Gamma-factor exceptional set and the pole handled exactly, and the functional
equation realized as a reflection symmetry. Second, an \emph{equivalence
lattice}: RH is proved equivalent --- in the kernel, axiom-clean --- to the
existence of any one of a family of abstract \emph{faithful carriers}
(Hodge-index, arithmetic-site, geometric-square-root, Deninger-dictionary,
moment-problem, spectral-realization, Morse-index-zero, Fredholm-squared, \dots),
each an interface that records \emph{what kind of object a proof must supply},
and each shipped with its own \emph{falsifier} (a single off-line zero refutes
it) and with \emph{non-supply} lemmas certifying that naive (finite, diagnostic,
fake) candidates do not instantiate it. Third, a suite of finite, fully proved
\emph{no-go certificates} that rule out entire classes of naive strategy ---
the Davenport--Heilbronn fake-family obstruction, the finite-carrier
obstruction, the failure of one-sided prime-local positivity, the
multiplicativity-visibility obstructions, a scattering-parity obstruction, and
the no-margin / resolution-wall obstruction. The novel contribution is not a
new theorem about $\zeta$: it is the machine-checking itself. The
RH ``wall'' --- known informally as a scattering of equivalences and folklore
obstructions --- becomes one axiom-clean, mutually consistent, kernel-enforced
formal artifact. \textbf{The library does not prove RH, and does not claim to:}
every load-bearing carrier hypothesis is left unproven, several are proved
\emph{equivalent} to RH, and the genuinely open input is named explicitly.
\end{abstract}

\maketitle

\section{Introduction}\label{sec:intro}

\subsection*{What is formalized, and what is not}
The Riemann Hypothesis asserts that the nontrivial zeros of the Riemann zeta
function lie on the line $\Re s = \tfrac12$. In \code{mathlib} it is the
proposition
\begin{center}
\code{RiemannHypothesis : $\forall$ s, riemannZeta s = 0 $\to$
($\lnot\exists$ n, s = -2*(n+1)) $\to$ s $\ne$ 1 $\to$ s.re = 1/2.}
\end{center}
This paper is \emph{not} a proof of RH. It is a description of a Lean~4 library
that formalizes the \emph{structure} of the problem: the equivalent forms RH can
take, and the obstructions that constrain any proof. We call this structure the
\emph{logical geometry of the wall}. Informally this geometry is well known ---
RH has dozens of published equivalent reformulations
\cite{broughan,conrey,bombieri}, and the function-field analogue (Weil, Deligne)
makes the shape of a hypothetical proof a recurring subject of speculation
\cite{weil,connes}. What has \emph{not} existed is a single machine-checked
artifact in which these reformulations are stated precisely against a common
formal kernel, proved equivalent to one another, and confronted with the
obstructions in the same logical universe. That artifact is the contribution
here.

\subsection*{Three layers}
The library, \code{jensen-ladder}, has three layers.

\begin{enumerate}
\item \textbf{The reduction spine} (\S\ref{sec:spine}). A handful of
unconditional, axiom-clean theorems carry \code{mathlib}'s
\code{RiemannHypothesis} to a clean working endpoint: the reality of the
\emph{regular} zeros of the completed zeta function $\Lambda$ written in the
symmetric variable $\Xifn(z)=\Lambda(\tfrac12 + iz)$. The functional equation
becomes evenness of $\Xifn$; the trivial-zero / Gamma-factor exceptional set and
the pole at $s=1$ are filtered exactly.

\item \textbf{The equivalence lattice} (\S\ref{sec:lattice}). RH is proved
equivalent, in the kernel, to the existence of any one of a family of abstract
\emph{faithful carriers}. Each carrier is a Lean \code{structure} whose
``faithful rows'' would suffice to prove RH (\code{riemannHypothesis\_of\_%
faithfulRows}); each ships with a falsifier (\code{not\_faithfulRows\_of\_%
nonrealRegularXiZero}); the carriers are proved mutually equivalent, forming a
lattice whose top is RH.

\item \textbf{The no-go certificates} (\S\ref{sec:nogo}). Finite, fully proved
theorems that exclude naive proof strategies: that a single Euler-product
coefficient law is the only separator (the Davenport--Heilbronn fake gate),
that no finite spectrum can host infinitely many zeros, that one-sided
prime-local positivity fails, that finite ``visibility'' cannot force
multiplicativity, that a scattering ratio cannot equal $\Xifn$, and that no
margin/floor mechanism survives (the resolution wall).
\end{enumerate}

\subsection*{The point of mechanizing this}
A skeptic may object that an equivalence ``RH $\iff$ (an object exists that
would prove RH)'' is logically shallow. The objection is partly correct and we
take it seriously throughout (\S\ref{sec:honest}). The value of the
mechanization is fourfold and is \emph{not} mathematical depth of any single
equivalence:
\begin{itemize}
\item \emph{Faithful bookkeeping.} The reduction targets are stated as precise
Lean types, not prose. ``What object would suffice'' becomes a checkable
\code{structure}; ``what is still open'' becomes a named, unproven hypothesis of
known status.
\item \emph{Kernel-enforced consistency.} The carriers are proved
\emph{equivalent to one another}, so the routes cannot secretly disagree about
what RH needs; the falsifier direction is uniform across all of them; the no-gos
compose with the lattice in the same logical universe.
\item \emph{The no-gos are genuine theorems.} Unlike the equivalences, the
no-go certificates are finite, self-contained, unconditional results --- a
machine-checked census of what cannot work.
\item \emph{Honest separation.} The library distinguishes, at the type level,
proved analytic content about $\zeta$ from abstract interfaces that merely
\emph{record} an open problem. Every module that consumes an unconstructed
object says so in its doc-string.
\end{itemize}

\subsection*{Relation to Theorem M}
A separate, self-contained result --- ``Theorem M'', the unconditional
hyperbolicity of the critical Laguerre family $\Psi_d$ \cite{theoremM} --- is a
proved \emph{input} to one route of the lattice (the model-to-$\Xifn$ transfer).
We emphasize, as that work does and as every relevant module here repeats:
\textbf{Theorem M is proven, but Theorem M does not prove RH by itself.} The
present library makes precise exactly which additional, unproven input would be
required.

\subsection*{Verification at a glance}
The toolchain is pinned (\code{leanprover/lean4:v4.30.0}, \code{mathlib}
\code{v4.30.0}). The complete check is
\begin{center}
\code{cd formal \&\& lake build}\qquad(green, $128$ modules, $8603$ jobs)
\end{center}
followed by \code{scripts/check\_axioms.sh}, which prints the axiom dependency of
a curated set of $16$ headline theorems spanning all three layers and confirms
each is exactly \code{[propext, Classical.choice, Quot.sound]} with zero
\code{sorry}s (\S\ref{sec:repro}).

\section{The reduction spine}\label{sec:spine}

The endpoint of the analytic program is most naturally stated about the
\emph{completed} zeta function $\Lambda=\code{completedRiemannZeta}$ and its
symmetric variable. Module \code{RHReduction} formalizes the bridge from
\code{mathlib}'s \code{RiemannHypothesis} to that endpoint, unconditionally.

\subsection*{The Gamma-factor bookkeeping}
\code{mathlib} provides $\zeta(s)=\Lambda(s)/\Gammar(s)$ for $s\neq 0$, where
$\Gammar(s)=\pi^{-s/2}\Gamma(s/2)$ vanishes exactly on $\{0,-2,-4,\dots\}$.
These are precisely the points excluded from the RH quantifier (the trivial
zeros, together with $s=0$, where $\zeta(0)=-\tfrac12\neq 0$). Hence at every
point in scope of the RH quantifier $\Gammar(s)\neq0$, so a zero of $\zeta$
there is a zero of $\Lambda$. Define a \emph{regular} zero to be one with
$\Gammar(s)\neq0$ and $s\neq1$. The module proves both directions and packages
them as an equivalence.

\begin{theorem}[\code{riemannHypothesis\_iff\_completedZeta\_regular\_zeros\_on\_line}]
\code{RiemannHypothesis} holds if and only if every regular zero $s$ of
$\Lambda$ has $\Re s=\tfrac12$.
\end{theorem}

\subsection*{The symmetric variable and the functional equation}
Set $\Xifn(z)=\Lambda(\tfrac12+iz)$ (\code{riemannXi}). The change of variables
$s=\tfrac12+iz$ sends $\Re s=\tfrac12$ to $\Im z=0$, and a regular zero of
$\Lambda$ to a \emph{regular $\Xifn$-zero} (\code{riemannXiRegularZero}: the
$\Xifn$-zero condition together with the two regularity filters). The functional
equation $\Lambda(1-s)=\Lambda(s)$ becomes the evenness of $\Xifn$.

\begin{theorem}[\code{riemannXi\_even}, \code{riemannXiRegularZero\_neg}]
$\Xifn(-z)=\Xifn(z)$ for all $z\in\C$, and the regular-zero predicate is
invariant under $z\mapsto -z$. The fixed-point set of this reflection is the
real axis $\Im z=0$, i.e.\ the critical line.
\end{theorem}

\noindent The second statement uses one nontrivial \code{mathlib} input
(nonvanishing of $\zeta$ on $\Re s\ge 1$) to keep the regularity filters
symmetric; it is the formal heart of ``functional equation $=$ reflection about
the critical line''. Composing the change of variables with the Gamma-factor
bridge yields the spine's headline equivalence.

\begin{theorem}[\code{riemannHypothesis\_iff\_regular\_riemannXi\_zeros\_real}]
\label{thm:spine}
\code{RiemannHypothesis} $\iff$ every regular $\Xifn$-zero is real
($\Im z=0$).
\end{theorem}

All of \S\ref{sec:spine} is unconditional and axiom-clean. It is the one place
in the library where ``RH'' is connected to a tractable analytic object; every
later layer targets the right-hand side of Theorem~\ref{thm:spine}.

\section{The equivalence lattice}\label{sec:lattice}

\subsection*{The carrier pattern}
Each route is organized around an abstract \code{structure} --- a
\emph{carrier} --- whose fields are the rows a proof along that route would have
to supply. The carrier is deliberately abstract: it does \emph{not} construct
the geometric or spectral object, only names the properties that object must
have. Every carrier module instantiates the same three-part pattern:
\begin{enumerate}
\item \textbf{sufficiency:} \code{riemannHypothesis\_of\_faithfulRows} --- the
faithful rows imply RH (via Theorem~\ref{thm:spine});
\item \textbf{falsifier:} \code{not\_faithfulRows\_of\_nonrealRegularXiZero} ---
a single off-line regular $\Xifn$-zero refutes the rows;
\item \textbf{non-supply:} lemmas of the form \code{X\_do\_not\_supply\_Y}
exhibiting an explicit \emph{diagnostic} carrier (e.g.\ empty spectrum, fake
coefficient stream) that satisfies the cheap rows but provably not the faithful
ones.
\end{enumerate}
The non-supply lemmas are what keep the pattern honest: they certify that the
carrier's faithful row is not vacuously satisfiable, so the equivalence does not
collapse to a triviality at the cheap end.

\subsection*{The lattice}
The carriers are then linked by kernel-checked \code{iff}s into a single lattice
whose top element is RH. Representative equivalences (all axiom-clean):
\[
\begin{aligned}
\text{RH}
&\iff \exists\ \text{a faithful Hodge-index carrier}\\
&\iff \exists\ \text{a faithful arithmetic-site carrier}\\
&\iff \exists\ \text{a faithful geometric-square-root carrier}\\
&\iff \exists\ \text{a faithful Fredholm-squared carrier}\\
&\iff \exists\ \text{a regular spectral realization}\\
&\iff \text{no negative Morse modes (Morse index $0$)},
\end{aligned}
\]
with the corresponding module and theorem names collected in
Table~\ref{tab:lattice}.
The intermediate equivalences are proved \emph{directly between carriers}
(e.g.\ \code{hasFaithfulHodgeIndexCarrier\_iff\_hasFaithfulArithmeticSiteCarrier},
\code{hasFaithfulGeometricSquareRootCarrier\_iff\_hasFaithfulHodgeIndexCarrier}),
so the lattice is internally consistent: every route agrees, in the kernel, on
what object RH requires. Table~\ref{tab:lattice} lists the carrier modules.

\begin{table}[t]
\scriptsize
\centering
\begin{tabular}{@{}p{0.42\textwidth}p{0.52\textwidth}@{}}
\toprule
Module & Carrier / equivalence (open content $=$ existence of the faithful rows)\\
\midrule
\code{HodgeIndexCarrier} & RH $\iff$ faithful Hodge-index carrier $\iff$ arithmetic-site carrier \\
\code{ArithmeticSiteCarrier} & RH $\iff$ faithful arithmetic-site carrier (Lefschetz/Weil-trace rows) \\
\code{GeometricSquareRootCarrier} & nonnegative Weil form $\iff$ abstract square root; $\iff$ Hodge-index carrier \\
\code{HodgeWeilBridge} & finite Hodge realization $\Rightarrow$ Weil form PSD; negative vector $\Rightarrow$ no realization \\
\code{PrimitiveHodgeWeilEngine/Calibration} & primitive Hodge--Riemann engine for finite Weil forms \\
\code{DeningerCarrier} & RH $\iff$ polarized faithful dictionary (flow / determinant interface) \\
\code{DeningerFlowUnitary}, \code{FriedDeningerTorsionCarrier} & flow-unitarity / torsion variants of the Deninger dictionary \\
\code{MomentProblemCarrier} & RH $\iff$ faithful moment-problem carrier $\iff$ spectral realization \\
\code{SpectralRealization} & RH $\iff$ $\exists$ (regular) spectral realization of the $\Xifn$-zeros \\
\code{SpectralReflection} & RH $\iff$ $\exists$ reflected (FE-symmetric) regular spectral realization \\
\code{ComplexSpectralDeterminantCarrier}, \code{UnitaryCokernelCarrier} & complex-determinant / unitary-cokernel realizations \\
\code{FredholmSquaredCarrier} & RH $\iff$ faithful Fredholm-squared carrier $\iff$ nonnegative squared support \\
\code{SquaredVariablePullback} & reality from nonnegative squared support / approximation \\
\code{PrimeCepstrumHankelCarrier} & prime-cepstrum Hankel realization (feeds Fredholm-squared) \\
\code{MorseCriterion} & RH $\iff$ no negative modes $\iff$ nonnegative spectral bottom \\
\code{MorseDeningerBridge} & no negative modes $\iff$ polarized faithful dictionary $\iff$ RH \\
\code{BochnerTopRoute} & RH $\iff$ nonpositive ``top'' (Bochner simple-top calibration) \\
\code{MagicInterpolationBridge} & magic-interpolation data $\Rightarrow$ RH; off-line zero $\Rightarrow$ no magic rows (sufficiency $+$ falsifier; Viazovska-style target) \\
\code{ModelToXiTransfer}, \code{PolyaSchurTransfer} & model endpoint (Theorem M) $+$ transfer row $+$ Jensen gate $\Rightarrow$ RH \\
\code{TruncationLimitCarrier}, \code{SpectralPollutionFreeLimit} & exhaustive / pollution-free limit carriers \\
\code{CVSSpectralRoute} (+\,\code{CVS*}, \code{CCMGroundState*}, \code{OddMorseCertificate}) & Connes--van Suijlekom finite-scale ground states $+$ convergence $\Rightarrow$ RH \\
\code{DeterminantHurwitzRoute}, \code{SquaredDeterminant*}, \code{DiagonalFredholmProduct} & \code{det\_reg}$\to\Xifn$ squared-determinant route \\
\bottomrule
\end{tabular}
\caption{The equivalence lattice. In every row the \emph{open content} is the
existence of the faithful rows; the equivalence and the falsifier are
proved.}\label{tab:lattice}
\end{table}

\subsection*{A worked carrier: the finite Hodge-to-Weil bridge}
To show the pattern carries genuine finite content (not only abstract
plumbing), consider \code{HodgeWeilBridge}. For a finite semilocal Weil matrix
$D$ (archimedean part minus prime/local part) it defines the quadratic form
$Q_D(v)=\sum_{i,j} v_i\,(\code{entry}\,D\,i\,j)\,v_j$ and proves the exact split
\[
Q_D(v) \;=\; Q_D^{\mathrm{arch}}(v) \;-\; \sum_{q} Q_D^{\mathrm{prime},q}(v)
\]
(\code{quadraticForm\_eq\_arch\_sub\_sum\_localPrimeQuadraticForm}). It then
defines an abstract \code{FiniteHodgeRealization} --- a map from finite
vectors to a \code{DivisorClass} with an intersection form, an abstract
\code{primitive} predicate, the Hodge-index row (primitive self-intersections
$\le 0$), and the identification $Q_D(v)=-\langle\text{realize }v,
\text{realize }v\rangle$ --- and proves that any such realization makes $Q_D$
positive semidefinite, while a single negative vector rules out every
realization. The doc-string states the boundary precisely: ``This file does not
construct the arithmetic surface, divisor classes, intersection theory, trace
identity, semilocal Connes operator, convergence theorem, or RH. It only records
the finite algebraic bridge that such a construction must instantiate.'' That is
the honest shape of the whole lattice: real finite algebra, an explicitly
abstract geometric interface, and a named open problem.

\section{The no-go certificates}\label{sec:nogo}

The second class of results is mathematically self-contained: finite,
unconditional theorems that rule out naive strategy classes. These are the
machine-checked ``wall''. Table~\ref{tab:nogo} lists them; we expand the two
most structurally important.

\subsection*{Multiplicativity is the only separator (Davenport--Heilbronn)}
A recurring failure mode is any argument that uses only the functional equation,
reality, and density/coefficient information --- because the
Davenport--Heilbronn function has all of those and yet has zeros off the
critical line. Module \code{DHMultiplicityFakeGate} formalizes the finite
arithmetic core that distinguishes $\zeta$ from such fakes: a single normalized
Euler-product coefficient stream must be (coprime) multiplicative, and the
Davenport--Heilbronn residue pattern already violates this at the coprime pair
$(2,3)$.

\begin{theorem}[\code{not\_coprimeMultiplicative\_of\_DHPattern236}]
If $a(1)=1$, $a(2)=\kappa$, $a(3)=-\kappa$, $a(6)=1$, then $a$ is not coprime
multiplicative --- multiplicativity would force $a(6)=a(2)a(3)=-\kappa^2\neq1$.
\end{theorem}

\noindent The companion modules \code{FiniteMultiplicityVisibilityNoGo},
\code{FiniteMixedLogVisibilityNoGo}, \code{FiniteEulerProductMixedLogGate}
sharpen this into a \emph{visibility} statement: a finite ``visible predicate''
(any finite window of coefficient data) cannot force coprime multiplicativity,
because a product-spike pattern agrees with the unit stream on the window yet
violates the law off it. This is the formal version of ``the Euler product is
the unique separator, and no finite amount of it suffices''.

\subsection*{No margin (the resolution wall)}
RH sits at an \emph{exact} threshold: empirically the minimal normalized
zero-gap tends to $0$ (the Lehmer phenomenon), so no proof can rely on a uniform
positive margin/floor. Module \code{ResolutionWall} (with \code{BAH1Quadratic})
formalizes the contradiction.

\begin{theorem}[\code{not\_forall\_certificate\_margin\_of\_uniform\_floor\_arbitrarily\_small\_margins}]
If certificate margins are bounded below by a fixed positive floor, but the
required margins are arbitrarily small, no certificate can meet the margin at
every index --- a contradiction. Hence a margin/floor mechanism cannot certify
the critical (marginless) inequality.
\end{theorem}

\noindent This is the formal counterpart of the analytic ``no-floor'' fact and
explains, inside the kernel, why soft-margin and single-positive-functional
methods are excluded. \code{CVSKatoSmallNoGo} reaches the same conclusion on the
spectral side: Kato-smallness fails precisely when the spectral gap collapses.

\begin{table}[t]
\scriptsize
\centering
\begin{tabular}{@{}p{0.42\textwidth}p{0.52\textwidth}@{}}
\toprule
Module & What is proved (finite, unconditional)\\
\midrule
\code{DHMultiplicityFakeGate} & the Davenport--Heilbronn pattern is not (coprime/totally) multiplicative \\
\code{FiniteMultiplicityVisibilityNoGo} & a finite visible predicate cannot force coprime multiplicativity \\
\code{FiniteMixedLogVisibilityNoGo}, \code{FiniteEulerProductMixedLogGate} & finite mixed-log data cannot force the product (Euler-coupling) relations \\
\code{FiniteCarrierNoGo} & a finite spectrum cannot represent infinitely many distinct zeros \\
\code{PrimeLocalNoGo} & one-sided prime-local kernel is negative below threshold $\Rightarrow$ local positivity fails \\
\code{PrimeLocalHodgeNoGo} & one-sided prime-local data admits a negative Hodge vector \\
\code{ScatteringParityNoGo} & a scattering ratio (even log-derivative) cannot equal $\Xifn$ (odd) --- closes the ratio class \\
\code{ResolutionWall}, \code{BAH1Quadratic} & uniform floor $+$ arbitrarily small margins $\Rightarrow$ contradiction (no margin) \\
\code{CVSKatoSmallNoGo} & Kato-smallness fails under critical gap collapse \\
\code{AdversarialSignTwistNoGo} & an adversarial sign-twist forces a negative direction unless budget $\le$ archimedean \\
\code{AbsoluteBudgetFakeFamilyBlindness} & absolute-budget certificates are blind to sign-retwisted fake families \\
\code{DeterminantNormalizationNoGo} & post-hoc scalar normalization cannot repair \code{det\_reg} \\
\code{SpectralFaithfulnessGap} & an off-axis regular zero $\Rightarrow$ the squared spectrum is not faithful \\
\code{SchurStarSharpness} & archimedean $<$ response budget $\Rightarrow$ an explicit negative form (sharpness) \\
\bottomrule
\end{tabular}
\caption{The no-go certificates: a machine-checked census of strategies that
cannot work.}\label{tab:nogo}
\end{table}

\section{Finite positivity algebra and unconditional analytic bricks}
\label{sec:bricks}

Two further families of modules carry genuine mathematical content below the
abstract interfaces.

\subsection*{Finite positivity algebra}
A toolkit of finite linear algebra realizes the ``square-root/domination''
mechanics the positivity routes need: \code{HodgeSignatureBabyCarrier},
\code{PrimeDominationBabyCarrier}, \code{GraphLaplacianBabyCarrier},
\code{RowDiagonalDominationCarrier}, \code{GlobalDominationReduction},
\code{SchurStarDominationCarrier}/\code{AssemblyCarrier}, and the finite-rank
\code{CCM*} algebra (\code{CCMFiniteWeil}, \code{CCMRankOne}, \code{CCMQuotient},
\code{CCMFrobeniusBound}, \code{CCMPerturbationBound}, \code{CCMFiniteKatoBridge}).
These prove, e.g., that an off-diagonal coupling dominated by the diagonal makes
a completed block positive semidefinite, and that a rooted incidence assembly's
energy is an exact square. The \code{Chiral*} modules formalize a finite
chiral-Dirac perfect-square route (PSD Hankel moments, Stieltjes trace
identities) with explicit ``diagnostic rows do not supply live rows'' honesty
lemmas.

\subsection*{Unconditional analytic bricks}
Several modules prove substantive, RH-independent facts about the completed zeta
function, some filling gaps not in \code{mathlib}:
\begin{itemize}
\item \code{RvMXiEntire} --- the entire completion $\code{xiE}(s)=\tfrac12\,
(s(s-1)\,\code{completedRiemannZeta}_0(s)+1)$ is entire, with nonnegative
divisor, and $\code{xiE}(s)=0 \iff \zeta(s)=0$ on the critical strip
(\code{xiE\_zero\_iff\_zeta\_zero}); the Riemann--von Mangoldt foundation.
\item \code{ZeroCountingN} --- the zero-counting function $N(T)$ as a finite,
monotone count.
\item \code{XiOrderBound}, \code{XiZeroCounting}, \code{W1DensityNevanlinna},
\code{CompletedZetaStripBound} --- order-$1$ growth bounds, a Jensen
zero-counting bound, vertical-strip bounds, and the Nevanlinna bricks toward
$\sum_\rho 1/(\tfrac14+\gamma^2)<\infty$.
\item \code{HurwitzRealRootedLimit} --- a \emph{from-scratch} Lean
formalization of the argument principle (zeros-in-a-contour) and the
nowhere-zero Hurwitz theorem, neither in \code{mathlib}, giving the headline
analytic reduction \code{det\_reg}$\to\Xifn\Rightarrow$RH used by the
squared-determinant route.
\end{itemize}
These are the genuinely deep unconditional content of the library and are of
independent formalization interest.

\section{Methodology and discipline}\label{sec:method}

\subsection*{Axiom hygiene}
The headline theorems of all three layers depend on exactly the three standard
\code{mathlib} axioms \code{[propext, Classical.choice, Quot.sound]}; the tree
contains no \code{sorry} and no added \code{axiom}. The script
\code{scripts/check\_axioms.sh} re-derives this for a curated set of $16$
headline names (Appendix~\ref{sec:repro}); it is wrap-tolerant (Lean line-wraps
long axiom lists) and fails loudly on any extra axiom, missing name, or
\code{sorry} token. As an independent safeguard, the exported headline
declarations and their full dependency closure ($54{,}693$ declarations) are
re-checked by Nanoda,\footnote{\url{https://github.com/ammkrn/nanoda_lib}; the
export uses \code{lean4export}, \url{https://github.com/leanprover/lean4export}.}
an independent Rust implementation of the Lean~4 kernel, with the permitted
axioms restricted to exactly the three above (\code{scripts/check\_nanoda.sh}).
Continuous integration runs the full \code{lake build}, the axiom audit, and the
Nanoda re-check on every push.

\subsection*{The falsifier discipline}
Every carrier ships with a falsifier: a single nonreal regular $\Xifn$-zero
refutes its faithful rows. This is not decoration --- it forces each interface
to be \emph{about} the zeros (an interface satisfiable independently of the zero
locations would have no falsifier), and it makes the lattice a genuine
reduction rather than a relabeling.

\subsection*{Abstract-interface honesty}
Where a module consumes an unconstructed object, its doc-string says so, and a
non-supply lemma exhibits a concrete cheap carrier that fails the faithful rows.
The library never imports an object it has not built; the open input is always a
named, unproven hypothesis.

\subsection*{Production process}
The development was produced by AI research agents (Claude ``Fable'' and GPT)
under the author's direction, in a disjoint-file-ownership mutual-audit
protocol, with the Lean kernel as final referee --- the same process used for
Theorem M \cite{theoremM}.

\section{Honest scope and limitations}\label{sec:honest}

We state the limitations plainly, because they determine what the artifact is
worth.
\begin{enumerate}
\item \textbf{It is not a proof of RH}, and contains none. Every carrier's
faithful rows are left unproven; no module instantiates a faithful carrier.
\item \textbf{The carrier equivalences are, by design, logically shallow.} A
carrier is essentially ``an object whose existence would prove RH and which a
single off-line zero refutes''; proving RH $\iff$ (such an object exists) is not
a deep theorem about $\zeta$. Their worth is bookkeeping, mutual consistency,
falsifier-faithfulness, and the non-supply lemmas --- not depth. We do not claim
otherwise.
\item \textbf{The genuinely substantive content is elsewhere:} the no-go
certificates (\S\ref{sec:nogo}), the from-scratch argument-principle/Hurwitz
analysis and the Riemann--von Mangoldt bricks (\S\ref{sec:bricks}), and the
reduction spine (\S\ref{sec:spine}).
\item \textbf{The single open input,} reached independently by the geometric,
spectral, moment, and Morse routes, is one object: a non-spectral geometric
carrier (an arithmetic $\Spec\Z\times_{\mathbb F_1}\Spec\Z$ / Deninger dynamical
cohomology / Connes--Consani arithmetic site) delivering a Hodge-index-positive
intersection form, equivalently a faithful zero dictionary. The library names
this object and its falsifier; it does not construct it. Constructing it is the
open problem.
\end{enumerate}

\section{Reproducibility}\label{sec:repro}

\subsection*{Build}
\begin{verbatim}
cd formal && lake build      # green: 128 modules, 8603 jobs
\end{verbatim}
Toolchain: \code{leanprover/lean4:v4.30.0}; \code{mathlib} pinned at
\code{v4.30.0} (see \code{lake-manifest.json}).

\subsection*{Axiom + sorry audit}
\begin{verbatim}
scripts/check_axioms.sh
  -> 16 headline theorems axiom-clean
     ([propext, Classical.choice, Quot.sound]), zero sorries.
\end{verbatim}
The audited headline names include the spine
(\code{riemannHypothesis\_iff\_regular\_riemannXi\_zeros\_real},
\code{riemannXi\_even}, \code{riemannXiRegularZero\_neg}), the lattice
(\code{hasFaithfulHodgeIndexCarrier\_iff\_riemannHypothesis} and its
carrier-to-carrier links, \code{FredholmSquaredCarrier},
\code{SpectralRealization}), the no-gos
(\code{not\_coprimeMultiplicative\_of\_DHPattern236},
\code{ScatteringParityNoGo}, \code{PrimeLocalNoGo}, \code{ResolutionWall}), and
an analytic brick (\code{xiE\_zero\_iff\_zeta\_zero}).

\subsection*{Sorry sweep}
\begin{verbatim}
# no real sorry/admit tokens anywhere in the tree:
rg -n '(^|[^A-Za-z_])(sorry|admit)([^A-Za-z_]|$)' \
   formal/JensenLadder
\end{verbatim}

\section{Related work}\label{sec:related}

\emph{Formalized analytic number theory.} Substantial parts of the analytic
theory of $\zeta$ are already formalized. In Isabelle/HOL, Eberl and Paulson
formalized a large portion of analytic number theory --- including an analytic
proof of the Prime Number Theorem and the Riemann/Hurwitz zeta and Dirichlet
$L$-functions \cite{eberl}. In Lean, the \code{PrimeNumberTheoremAnd} project
(Kontorovich, Tao, and collaborators) formalized the Prime Number Theorem with
error term \cite{pnta}, and \code{mathlib} itself supplies the \emph{statement}
\code{RiemannHypothesis} \cite{mathlib} that this library targets. All of these
formalize \emph{theorems about} $\zeta$; none addresses RH's equivalent
reformulations or the obstructions to proving it. The present work is
complementary --- it mechanizes the \emph{logical geometry} around the open
statement, not an analytic theorem implying it.

\emph{The informal RH landscape.} RH has an extensive informal catalogue of
equivalent reformulations --- two volumes by Broughan \cite{broughan}, Conrey's
survey \cite{conrey}, and the official problem description \cite{bombieri} ---
while the function-field analogue (Weil \cite{weil}, Deligne) and the
noncommutative-geometry program (Connes \cite{connes}) supply the
``geometric positivity'' target that the carriers of \S\ref{sec:lattice}
abstract. To our knowledge none of this had previously been mechanized; the
contribution here is to state these equivalences and obstructions precisely
against one kernel, prove them mutually consistent, and pair each carrier with a
falsifier and each obstruction with a finite certificate.

\section{Conclusion}\label{sec:concl}

The Riemann Hypothesis has long had a rich informal ``geometry'': a web of
equivalent reformulations and a folklore of obstructions explaining why the
easy routes fail. \code{jensen-ladder} turns that geometry into a single
machine-checked object. The reformulations become a kernel-enforced equivalence
lattice; the obstructions become finite, unconditional no-go certificates; the
boundary between proved analytic content and abstract reduction target is made
explicit at the type level; and the one open input is named, not hidden. None of
this proves RH --- and the library is scrupulous never to suggest it does. What
it offers is a faithful, axiom-clean map of where the problem actually sits, of
a kind that, to our knowledge, has not previously been mechanized.

\subsection*{Acknowledgments}
The library and its formal verification were developed by AI research agents,
Claude (``Fable'', Anthropic) and GPT (OpenAI), under the author's direction,
with mutual line-audits and the Lean kernel as referee.

% Bibliography verified against primary sources (2026-06).
\begin{thebibliography}{99}

\bibitem{lean4}
L.~de~Moura and S.~Ullrich,
\emph{The Lean 4 theorem prover and programming language},
in: Automated Deduction --- CADE 28, LNCS \textbf{12699}, Springer, 2021,
625--635. DOI: \url{https://doi.org/10.1007/978-3-030-79876-5_37}.

\bibitem{mathlib}
The mathlib Community,
\emph{The Lean mathematical library},
in: Proc.\ 9th ACM SIGPLAN Int.\ Conf.\ on Certified Programs and Proofs
(CPP 2020), ACM, 2020, 367--381. DOI: \url{https://doi.org/10.1145/3372885.3373824}.

\bibitem{eberl}
M.~Eberl,
\emph{Nine Chapters of Analytic Number Theory in Isabelle/HOL},
in: 10th Int.\ Conf.\ on Interactive Theorem Proving (ITP 2019),
LIPIcs \textbf{141}, Schloss Dagstuhl, 2019, Art.~16.

\bibitem{pnta}
A.~Kontorovich, T.~Tao, et al.,
\emph{PrimeNumberTheoremAnd}, Lean~4 formalization of the Prime Number Theorem,
2024. \url{https://github.com/AlexKontorovich/PrimeNumberTheoremAnd}.

\bibitem{theoremM}
M.~Haufschild,
\emph{Unconditional hyperbolicity of the critical Laguerre family: an
elementary and formally verified proof}, preprint, 2026.
% companion repository: theorem-m

\bibitem{broughan}
K.~Broughan,
\emph{Equivalents of the Riemann Hypothesis. Vol.~1: Arithmetic Equivalents;
Vol.~2: Analytic Equivalents}, Encyclopedia of Mathematics and its
Applications \textbf{164}--\textbf{165}, Cambridge Univ.\ Press, 2017.

\bibitem{conrey}
J.~B. Conrey,
\emph{The Riemann Hypothesis},
Notices Amer.\ Math.\ Soc.\ \textbf{50} (2003), no.\ 3, 341--353.

\bibitem{bombieri}
E.~Bombieri,
\emph{Problems of the Millennium: The Riemann Hypothesis},
Clay Mathematics Institute, 2000.

\bibitem{weil}
A.~Weil,
\emph{Sur les courbes alg\'ebriques et les vari\'et\'es qui s'en d\'eduisent},
Actualit\'es Sci.\ Ind.\ \textbf{1041}, Hermann, Paris, 1948.

\bibitem{connes}
A.~Connes,
\emph{Trace formula in noncommutative geometry and the zeros of the Riemann
zeta function}, Selecta Math.\ (N.S.) \textbf{5} (1999), no.\ 1, 29--106.
DOI: \url{https://doi.org/10.1007/s000290050042}.

\end{thebibliography}

\end{document}
